A best practices composite lifetable for the United States

### 2016-02-28 BPLE/Manuscript.tex
this project needs to be kickstarted again for upcoming lab talk. Soon to become focus

\usepackage{appendix}
\usepackage{amsmath}
\usepackage{caption}
\usepackage{placeins}
\usepackage{graphicx}
\usepackage{subcaption}
%\usepackage{subfig}
\usepackage{longtable}
\usepackage{setspace}
%\usepackage{tikz}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{xcolor,colortbl}
\usepackage{chngpage}
%\usepackage[active,tightpage]{preview}
\usepackage{natbib}
\bibpunct{(}{)}{,}{a}{}{;} 
\usepackage{url}
\usepackage{nth}
\usepackage{authblk}
\usepackage[most]{tcolorbox}
%\usepackage{hyperref}
%\usepackage{color}
%\usepackage{fontspec}
%\usepackage{pdfsync}
\usepackage[normalem]{ulem}
\usepackage{amsfonts}
\renewcommand{\listtablename}{List of Appendix Tables}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
% working on this need to concatenate file name based on sex and variable name
%\newcommand\Cell[1]{{\raisebox{-0.05in}{\includegraphics[height=.2in,width=.2in]{Figures/ColorCodes/\expandafter#1}}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% setting color to letters affects spacing. Here's a hack I found here:
% http://tex.stackexchange.com/questions/212736/change-letter-colour-without-losing-letter-spacing
%\DeclareRobustCommand{\spacedallcaps}[1]{\MakeUppercase{\textsc{#1}}} % all
% caps with better spacing

@@

\defcitealias{HMD}{HMD}

@@

\begin{abstract}
We calculate a preliminary series of best-possible lifetables for the United
States from 1959 to 2004, defined on the basis of the age-specific aggregate of
the lowest observed age-cause specific death rates among the 50
states and the District of Columbia for each year. The resulting highest-possible life expectancy shows a gradual increasing trend over the period, on
average 1.9 and 2.2 years higher than the highest state life expectancy in each year for males and females, respectively. We argue that the US states best practices lifetable is a useful
gauge of future mortality trends and a good benchmark against which to compare
state mortality patterns.
\ackn{The work reported in this manuscript was supported in part by the U.S.
National Institute On Aging of the National Institutes of Health under award
numbers R01-AG011552 and R01-AG040245. The content is solely the responsibility of the authors and does not necessarily represent the official views of the funding agencies.}
\end{abstract}

@@

The level and timing of cause-specific responses to particular
technological or well-being improvements varies between populations. Therefore,
the lowest observed all-cause force of mortality is not necessarily composed of
the lowest cause-specific forces of mortality, which means that the vanguard life
expectancy may not reflect the best mortality possible given current conditions
over a set of populations. For instance, the lowest rate of suicide and heart
disease at age 45 will not necessarily be found in the same population. As an
alternative to vanguard life expectancy, one could create cause-specific
low-mortality benchmarks, or more synthetically, combine the lowest mortality
observed by age and cause into a hybrid minimum mortality schedule. This synthetic low mortality lifetable--- hybrid with respect to reference population over age and cause--- was already suggested by \citet{wunsch1975minimum} and \citet{vallin2008minimum}. These authors investigated trends at the national-level. We refer to such hybrid minimum
lifetables as ``best practices'' lifetables. \citet{eikemo2014} use the term
``best practices'' in a similar sense of an ideal mix of conditions, but do so
via combining particular risk factors with their estimated mortality impacts to
produce best case scenarios. The end effect is quite similar, to the extent that
either estimation strategy shows what society could achieve given perfect
diffusion of some definition of best conditions. In the case of a best practices
lifetable, we condition on mortality outcomes only, making the endeavor much
simpler in practice.

@@

\section*{Data and methods}
All death count data come from the public NCHS mortality microdata files
\citep{NCHSdata}.\footnote{The current analysis is based on the public files
with geographic information only up to 2004. We have bene granted access to
geographic variables in a NCHS Research Data Center, and will extend analysis
through 2013 for the final paper version.} Population denominators come from a
beta version of the U.S.
States project of the Human Mortality Database \citep{HMD}. Further processing of
population counts into exposures is done according to the HMD Methods
Protocol, and all-cause lifetables were calculated by single ages \citep{HMDMP}.
Death counts for causes were aggregated into quinquennial age groups by year and sex for eleven large groups that minimize disruptions between ICD transitions, and then
converted to fractions. Fractions were then multiplied into the single-age
all-cause death rates. The eleven causes of death considered in this abstract
version are displayed in Table~\ref{tab:desc}, with their relative percentages.

@@

One caveat of calculating a best practices lifetable is that stochastic zeros
occur in particular ages and causes, especially among small populations. The
exercise of selecting the minimum death rate for a given age and cause will tend
to collect stochastic zeros, producing an excessively optimistic best practices
life expectancy. We therefore smooth all death rates using
the \texttt{MortalitySmooth} package \citep{GC2012} in \texttt{R}. The age-period
matrix of death rates for each state and sex is smoothed over both margins
using a p-spline. Cause fractions are then recalculated, and
multiplied into the smoothed all-cause death rates. Finally, we calculate
lifetables for each sex and state, as well as the best practices lifetable of
each year using HMD methodology.

We decompose differences in state life expectancy from the best-practices life
expectancy using the pseudo-continuous method proposed by
\citet{horiuchi2008decomposition}. We prefer this method because it allows for
defining life expectancy as a direct function of age-cause-specific rates,
$M_{(x,c)}$, thereby eliminating some artifacts present in other methods.
Results are summarized by ages, causes and states and can be interpreted
directly as the the total lag in years of life expectancy due to excess
mortality (anything greater than the minimum) in the given margin.\footnote{In
the final version we may opt to further separate trends from initial
differences using the decomposition method recently proposed by
\citet{DimaDecomp2014}.}

\subsection*{Vanguard and best practices life expectancy}
For the sake of clarity, we define vanguard life expectancy as the highest
observed state life expectancy for each sex in a given year. Elsewhere, this
quantity is called the record life expectancy \citep{oeppen2002broken}, or even
the best practices life expectancy \citep{sanderson2004putting}, but we apply
the term best practices in a different sense. To define the best practices
mortality schedule, for each year and sex separately select the lowest
(smoothed) death rate by age, $x$, and cause, $c$, from among all the states, $s$, $m(s,x,c)$, and sum over
all causes, $C$, within ages to define the minimum mortality schedule,
$m(x)^{min}$
\begin{equation}
m(x)^{min} = \sum _{c=1}^C min(m(s,x,c))
\end{equation}

The minimum mortality schedule is then used to calculate the best-practices life
expectancy, and its cause-specific mortality pattern is used as the
benchmark against which to decompose the mortality pattern of each state. While
the best practices life expectancy is drawn from the diverse mortality
experience of the entire USA, the vanguard life expectancy refers only to a
particular state in a given year. We can say, for instance, that Minnesota was
the vanguard for females in the year 1959, with a lif expectancy of 75.3, but
that the best practices life expectancy in the same year was another 3 years
higher at 78.3.

There are limitations to the interpretation of a best practices trend, as
defined here. The mix of causes within each age, and the resulting age pattern
for each cause may not necessarily reflect a possible or likely mortality
pattern. For instance, diseases may compensate each other, especially for
diagnostic and reporting reasons. For instance, for decedents that suffered
from both condition A and condition B, one state might be more likely to
declare A as the underlying cause of death, while another state would more
likely declare B to be as the underlying cause. A situation like this
would tend to inflate the best practices $e(0)$. The best practices $e(0)$ may
be underestimate in instances where reductions in cause A are expected to lead
to reductions in cause B (e.g., pneumonia and cardiovascular disease.)
\FloatBarrier
\section*{Preliminary results}
\FloatBarrier

@@

For both males and females, the best
practices trend is on average 2 years higher than the vanguard trend, and its
progress is steadier than either the vanguard, national, or average state
trends, as well as most individual states. The best practices trend has increased each year for
males and females, though faster for males. The average annual improvement for
males has been 0.1790 years of $e(0)$ per annum, compared with 0.1785 for the
male national average life expectancy. The respective rates of improvement for
females have been 0.1470 and 0.1456. It is convenient and telling that the
respective medium term average rates of improvement between national and
best practices life expectancy match so closely, despite the best practices
trend having a much more stable trajectory. This artifact may provide some
support to conceiving of the best practices trend as indicative of where the
U.S. is going in the coming decades.

@@

If we take the mean rate of improvement for national or best practice trends,
and compare it with the means of each series, we can roughly conclude that the
best practices trend in year $t$ is a decent estimate of the national life
expectancy in 30 or so years. This time lag is valid for the present series,
but will vary depending on the number of cause groups that enter into the
selection of minimum mortality rates. Clearly, if the population units were
counties rather than states, the level of best practices life expectancy would
increase even further, but we opine that cause-specific smoothing at the county
level is more of a modelling gesture than sound practice, due to very low counts of deaths in
most counties in the U.S. In the final paper we will aim to test the cause set
proposed by \citet{vallin2008minimum} and compare it with other reasonable
configurations for the U.S., and so recommend a suitable cause grouping.

For the sake of comparison with the presently calculated best practices series.
The $e(0)$ value in 2004 for males and females was 80 and 84.9, respectively. In
the same year, the observed all-cause $e(0)$ in Japan were 78.6 for males and
85.5 for females. That is, the U.S. best practices trend is ahead of observed
life expectancy for Japanese males, and behind for Japanese females. We may also
compare with the official $e(0)$ projections of the U.S. Social Security
Administration (SSA). The SSA projects projects male $e(0)$ passing 80 in the
year 2058, and female $e(0)$
passing 84.9 in the year 2075 \citep{SSA2005}. These projections are a full 54
and 71 years after 2004, in both cases much farther in the future than a simple extrapolation
of the best practices trend would predict, circa 2034.

>>>>

\usepackage{appendix}
\usepackage{amsmath}
\usepackage{caption}
\usepackage{placeins}
\usepackage{graphicx}
\usepackage{subcaption}
%\usepackage{subfig}
\usepackage{longtable}
\usepackage{setspace}
%\usepackage{tikz}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{xcolor,colortbl}
\usepackage{chngpage}
%\usepackage[active,tightpage]{preview}
\usepackage{natbib}
\bibpunct{(}{)}{,}{a}{}{;} 
\usepackage{url}
\usepackage{nth}
\usepackage{authblk}
\usepackage[most]{tcolorbox}
%\usepackage{hyperref}
%\usepackage{color}
%\usepackage{fontspec}
%\usepackage{pdfsync}
\usepackage[normalem]{ulem}
\usepackage{amsfonts}
\renewcommand{\listtablename}{List of Appendix Tables}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
% working on this need to concatenate file name based on sex and variable name
%\newcommand\Cell[1]{{\raisebox{-0.05in}{\includegraphics[height=.2in,width=.2in]{Figures/ColorCodes/\expandafter#1}}}}  
\defcitealias{HMD}{HMD}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% setting color to letters affects spacing. Here's a hack I found here:
% http://tex.stackexchange.com/questions/212736/change-letter-colour-without-losing-letter-spacing
%\DeclareRobustCommand{\spacedallcaps}[1]{\MakeUppercase{\textsc{#1}}} % all
% caps with better spacing

@@

%\defcitealias{HMD}{HMD}

@@

\begin{abstract}
We calculate a preliminary series of best-possible lifetables for the United
States from 1959 to 2004, defined on the basis of the age-specific aggregate of
the lowest observed age-cause specific death rates among the 50
states and the District of Columbia. The resulting highest-possible life expectancy shows a gradual increasing trend over the period, on
average 1.9 and 2.2 years higher than the highest state life expectancy in each year for males and females, respectively. We argue that the US states best practices lifetable is a useful
gauge of future mortality trends and a good benchmark against which to compare
state mortality patterns.
\ackn{The work reported in this manuscript was supported in part by the U.S.
National Institute On Aging of the National Institutes of Health under award
numbers R01-AG011552 and R01-AG040245. The content is solely the responsibility of the authors and does not necessarily represent the official views of the funding agencies.}
\end{abstract}

@@

The level and timing of cause-specific responses to particular
technological or well-being improvements varies between populations. Therefore,
the lowest observed all-cause force of mortality is not necessarily composed of
the lowest cause-specific forces of mortality, which means that the vanguard life
expectancy may not reflect the best mortality possible given current conditions
in a set of populations. For instance, the lowest rate of suicide and heart
disease at age 45 will not necessarily be found in the same population. As an
alternative to vanguard life expectancy, one could create cause-specific
low-mortality benchmarks, or more synthetically, combine the lowest mortality
observed by age and cause into a hybrid minimum mortality schedule. This synthetic low mortality lifetable--- hybrid with respect to reference population over age and cause--- was already suggested by \citet{wunsch1975minimum} and \citet{vallin2008minimum}. These authors investigated trends at the national-level. We refer to such hybrid minimum
lifetables as ``best practices'' lifetables. \citet{eikemo2014} use the term
``best practices'' in a similar sense of an ideal mix of conditions, but do so
via combining particular risk factors with their estimated mortality impacts to
produce best case scenarios. The end effect is quite similar, to the extent that
either estimation strategy shows what society could achieve given perfect
diffusion of some definition of best conditions. In the case of a best practices
lifetable, we condition on mortality outcomes only, making the endeavor much
simpler in practice.

@@

\section*{Data and methods}
All death count data come from the public NCHS mortality microdata files
\citep{NCHSdata}.\footnote{The current analysis is based on the public files
with geographic information only up to 2004. We have bene granted access to
geographic variables in a NCHS Research Data Center, and will extend analysis
through 2013 for the final paper version.} Population denominators come from a
beta version of the U.S.
States project of the \citet{HMD}. Further processing of
population counts into exposures is done according to the HMD Methods
Protocol, and all-cause lifetables were calculated by single ages \citep{HMDMP}.
Death counts for causes were aggregated into quinquennial age groups by year and sex for eleven large groups that minimize disruptions between ICD transitions, and then
converted to fractions. Fractions were then multiplied into the single-age
all-cause death rates. The eleven causes of death considered in this abstract
version are displayed in Table~\ref{tab:desc}, with their relative percentages.

@@

One caveat of calculating a best practices lifetable is that stochastic zeros
occur in particular ages and causes, especially in small populations. The
exercise of selecting the minimum death rate for a given age and cause will tend
to collect stochastic zeros, producing an excessively optimistic best practices
life expectancy. We therefore smooth all death rates using
the method proposed by \citep{GC2012}. The age-period
matrix of death rates for each state and sex is smoothed over both margins
using a p-spline. Cause fractions are then recalculated, and
multiplied into the smoothed all-cause death rates. Finally, we calculate
lifetables for each sex and state, as well as the best practices lifetable of
each year using HMD methodology.

We decompose differences in state life expectancy from the best-practices life
expectancy using the pseudo-continuous method proposed by
\citet{horiuchi2008decomposition}. We prefer this method because it allows for
defining life expectancy as a direct function of age-cause-specific rates,
$M_{(x,c)}$, thereby eliminating some artifacts present in other methods.
Results are summarized by ages, causes and states and can be interpreted
directly as the the total gap in years of life expectancy due to excess
mortality (anything greater than the minimum) in the given margin.
%\footnote{In
%the final version we may opt to further separate trends from initial
%differences using the decomposition method recently proposed by
%\citet{DimaDecomp2014}.}

\subsection*{Vanguard and best practices life expectancy}
We define vanguard life expectancy as the highest
observed state life expectancy for each sex in a given year. Elsewhere, this
quantity is called the record life expectancy \citep{oeppen2002broken}, or even
the best practices life expectancy \citep{sanderson2004putting}, but we apply
the term best practices in a different sense. To define the best practices
mortality schedule, for each year and sex separately select the lowest
(smoothed) death rate by age, $x$, and cause, $c$, from among all the states, $s$, $m(s,x,c)$, and sum over
all causes, $C$, within ages to define the minimum mortality schedule,
$m(x)^{min}$
\begin{equation}
m(x)^{min} = \sum _{c=1}^C min(m(s,x,c))
\end{equation}

The minimum mortality schedule is then used to calculate the best-practices life
expectancy, and its cause-specific mortality pattern is used as the
benchmark against which to decompose the mortality pattern of each state. While
the best practices life expectancy is drawn from the diverse mortality
experience of the entire USA, the vanguard life expectancy refers only to a
particular state in a given year. We can say, for instance, that Minnesota was
the vanguard for females in the year 1959, with a life expectancy of 75.3, but
that the best practices life expectancy in the same year was 3 years
higher at 78.3.

There are limitations to the interpretation of a best practices trend, as
defined here. The mix of causes within each age, and the resulting age pattern
for each cause may not necessarily reflect a possible or likely mortality
pattern. Diseases may compensate each other, especially for
diagnostic and reporting reasons. For instance, for decedents that suffered
from both condition A and condition B, one state might be more likely to
declare A as the underlying cause of death, while another state would on
average more likely declare B to be as the underlying cause. A situation like
this would tend to inflate the best practices $e(0)$. The best practices $e(0)$ may
be underestimated in instances where reductions in cause A are expected to lead
to reductions in cause B (e.g., pneumonia and cardiovascular disease.)
\FloatBarrier
\section*{Preliminary results}
\FloatBarrier

@@

For both males and females, the best
practices trend is on average 2 years higher than the vanguard trend, and its
progress is steadier than either the vanguard, national, or average state
trends, as well as most individual states. The best practices trend has increased each year for
males and females, though faster for males. The average annual improvement for
males has been 0.1790 years of $e(0)$ per annum, compared with 0.1785 for the
male national average life expectancy. The respective rates of improvement for
females have been 0.1470 and 0.1456. It is convenient and revealing that the
respective medium term average rates of improvement between national and
best practices life expectancy match so closely, despite the best practices
trend having a much more stable trajectory. This artifact may provide some
support to conceiving of the best practices trend as indicative of where the
U.S. is going in the coming decades.

@@

If we take the mean rate of improvement for national or best practice trends,
and compare it with the means of each series, we can roughly conclude that the
best practices trend in year $t$ is a decent estimate of the national life
expectancy in 30 or so years. This time lag is valid for the present series,
but will vary depending on the number of cause groups that enter into the
selection of minimum mortality rates. If the population units were
counties rather than states, the level of best practices life expectancy would
increase even further, but this exercise would be more challenging, due to very
low counts of deaths in most counties in the U.S. Increasing the number of
causes of death would similarly increase estimates would increase in a like
manner, as would subdividing the population in other categories, such as race
or income. In the final paper we will aim to test the cause set proposed by
\citet{vallin2008minimum} and compare it with other reasonable configurations for the U.S., and so recommend a suitable cause grouping.

For the sake of comparison with the presently calculated best practices series.
The $e(0)$ value in 2004 for males and females was 80 and 84.9, respectively. In
the same year, the observed all-cause $e(0)$ in Japan were 78.6 for males and
85.5 for females. That is, the U.S. best practices trend is ahead of observed
life expectancy for Japanese males, and behind for Japanese females. We may also
compare with the official $e(0)$ projections of the U.S. Social Security
Administration (SSA). The SSA projects projects male $e(0)$ passing 80 in the
year 2058, and female $e(0)$
passing 84.9 in the year 2075 \citep{SSA2005}. These projections are a full 54
and 71 years after 2004, in both cases much farther in the future than a simple extrapolation
of the best practices trend would predict, around 2034. If this seems unlikely,
recall that the rate components necessary to acheive this improvement are
already observed in the United States!

Note that both male and female best practice trends increased unabated through
the 1960s, where life expectancy tended to stagnate at the national and state
levels (cit). Had we forecast mortality in the following decades based on data
only through the 1960s history tells us we would underestimate future life
expectancy. Even in the 1960s the best practices trend would have told
us to aim forecasts higher.
